---
output: github_document
bibliography: inst/REFERENCES.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# sars

> *fit and compare species-area relationship models using multimodel inference*

**sars** provides functionality to fit twenty SAR model using non-linear regression, and to calculate multi-model averaged curves using various information criteria. The software also provides easy to use functionality to plot multi-model SAR curves and to generate confidence intervals using bootstrapping.

## Installation

You can install the released version of sars from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("sars")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("txm676/sars")
```
## Example

Basic usage of sars will result in using to types of functions:

```{r include = FALSE}
library(sars)
```

Fitting the power sar model [@Arrhenius1921] to the 'galapagos' data set
```{r}
fit_pow <- sar_power(data = galap)
```

Attempting to fit all 20 sar models to the 'galapagos' data set and generate a multimodel inference
```{r warning = FALSE}
mm_galap <- multi_sars(data = galap)

```

Most of 'fitting' functions have corresponding plot methods.

Fitting the exponential sar model [@Gleason1922] to the 'galapagos' data set and plot it
```{r}
fit_expo <- sar_expo(data = galap)

plot(fit_expo)
```

## Troubleshoutting :bomb:

If, despite the :heart: brought during the programming of this R :package: and writing of this documentation, you have difficulties to install or run sars, if you have questions about the procedures or calculations, or if you want to report bugs :beetle:, do not hesitate to connect with us on [GitHub](https://github.com/txm676/sars).

# References
