---
title: "sars R Package"
author: Thomas J. Matthews and Francois Guilhaumon
output: rmarkdown::html_vignette
bibliography: REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{sars R Package}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%"
)
```



This vignette is heavily based on the paper that accompanies the package.To 
cite this vignette, please cite the corresponding paper:

Matthews, T.J., Triantis, K., Whittaker, R.J. and Guilhaumon, F. (2019) sars:
an R package for fitting, evaluating and comparing species–area relationship
models. Ecography, In press

## BACKGROUND 
The species–area relationship (SAR) describes the near universally observed
pattern whereby the number of species increases with the area sampled, and it
has been described as one of ecology’s few laws (@Rosenzweig1995). The SAR is
a fundamental component of numerous ecological and biogeographical theories,
such as the equilibrium theory of island biogeography (@MacArthurWilson1967).
In addition, SAR models have been widely used in applied ecology and
conservation biogeography: for example, to predict the number of extinctions
due to habitat loss. Numerous types of SAR have been described, and one
primary dichotomy employed is the split of SARs into island SARs (ISARs),
whereby each data point is an individual island or isolated sample, and
species accumulation curves (SACs) that represent cumulative counts of
increased species number with sampling area (@Gray2004). Whilst the remainder
of the paper and the described R package are focused on ISARs, the models and
the model fitting procedure can equally be applied to SACs (see
@Matthews2016a), although it should be noted that in SACs the data points are
not independent of one another.

Over 20 SAR models have been described in the literature (@Dengler2009,
@Tjorve2003, @Tjorve2009, @Triantis2012). However, despite this wide range of
models, the majority of SAR studies are still based exclusively on the power
model (@Arrhenius1921), which if fitted in its non-linear (untransformed)
form generally takes a convex form. Often, the log–log representation of the
power model is used as it can be fitted using standard linear regression, and
its parameters are more easily interpretable (@Rosenzweig1995). However,
whilst the power model has been found to provide a reasonable fit to a wide
range of datasets (@Dengler2009, @Matthews2016b), it is not universally the
best model, and a number of studies have reported other models to provide
better fits to empirical data (e.g. @Triantis2012, @Matthews2016b). The
possibility of scale dependency of the form of the SAR has long been of
interest, with, for example, a theoretical case being made for SARs at
intermediate spatial scales being approximated by a power model, whilst at
larger spatial scales the form of the SAR has been theorised to be sigmoidal.
Additionally, it is only recently that the SAR for archipelagos as units of
analysis and not just islands has started to be studied, and thus we know
little about the form of archipelago SARs.

Due to the increased recognition of model uncertainty in SAR research, a
number of recent studies have employed a multi-model inference approach
(@BurnhamAnderson2002) in the analysis of SARs, whereby either (1) multiple
SAR models are compared using various criteria (e.g. AIC) and a best model is
chosen (e.g. @Dengler2009), or (2) multiple SAR models are fitted and a
multi-model averaged curve is calculated using, for example, AIC weights
(e.g. @Guilhaumon2008). We are not aware of any published software package
that enables users to fit, and create multi-model averaged curves using more
than eight SAR models. Considering currently available software, the BAT R
package provides functions to fit three SAR models (linear, power and
logarithmic); however, this package is focused on general biodiversity
assessment and thus does not provide any additional SAR functionality. The
mmSAR R package (@Guilhaumon2010) is focussed on SARs and while it allows
users to fit eight SAR models using an information theoretic framework, it
does not include several models that have been found to provide the best fits
to several empirical datasets. To provide a set of tools to fill these gaps,
we have developed the R package ‘sars’. The package provides functions to fit
20 SAR models using non-linear and linear regression, calculate multi-model
averaged curves using various information criteria, and generate confidence
intervals using bootstrapping. Novel features compared with mmSAR include (i)
user-friendly functions  for plotting (the user can now plot weighted
multimodel SAR curves along with the individual SAR model curves) and (ii)
determining the observed shape of the model fit (i.e. linear, convex up,
convex down or sigmoidal) and (iii) presence or not of an asymptote, and (iv)
functions to fit, plot and evaluate

@Coleman1981's random placement model using a species-site abundance matrix,
and (v) to fit the general dynamic model (GDM) of island biogeography
(@Whittaker2008). In addition, the mmSAR package (which has been deprecated)
no longer complies with recognised programming good practice , is not on CRAN
(the main repository of R packages), and is not user friendly (e.g. it
requires the user to load individual models prior to fitting). There was
therefore a need to design a new package from scratch.

##METHODS AND FEATURES 
The ‘sars’ (species–area relationships) package has been programmed using
standard S3 methods and is available on CRAN (version 1.1.0) and the
development version on GitHub (txm676/sars), meaning researchers can easily
add in their own models and functions and integrate these into the
multi-model inference framework. Thus, the package represents a resource for
future SAR work that can be built on and expanded by workers in the field.

Fitting individual SAR models and a set of SAR models The package provides
functions to fit each of the 20 SAR models (see also
@Triantis2012). The ‘sar_models’ function can be used to bring up a list of
the 20 model names and a Table can be generated in the package using the
‘display_sars_models’ function. With the exception of the linear model (which
is fitted using standard linear regression), all models are fitted using
non-linear regression and the model parameters are estimated by minimizing
the residual sum of squares with an unconstrained Nelder-Mead optimization
algorithm and the ‘optim’ R function. The starting values for the parameter
estimates are carefully chosen to avoid numerical problems and to speed up
the convergence process. However, custom starting values can be provided for
any of the 20 models using the ‘start’ argument in the model fit functions.

Each individual model fit returns an object of class ‘sars’, which is a list
of 22 elements containing relevant model fit information, such as the model
parameter estimates, the fitted values, the residuals, model fit statistics
(e.g. AIC, R2), the observed model shape (linear, convex or sigmoidal),
whether or not the fit is asymptotic, and convergence information. The
returned object can easily be plotted using the ‘plot.sars’ generic function;
as this function is based on the base R plotting framework, the plot
aesthetics can be edited using standard plotting arguments (see the
‘plot.sars’ documentation in the package). Summary and print generic
functions are also provided for class ‘sars’; these functions follow the
output of the standard ‘lm’ function in the ‘stats’ R package.
Multiple SAR models can be fitted to the same dataset using the ‘sar_multi’
function and the resultant n model fit objects stored together as a
‘fit_collection’ object. This object is a list of class ‘sars’, where each of
the n elements contains an individual SAR model fit. Using the ‘plot.sars’
generic function on a ‘fit_collection’ object generates a grid of the n
individual model fit plots (Fig. 1).

```{r include = FALSE}
library(sars)
```

```{r, fig.width=6, fig.height=6}
#load an example dataset (Preston, 1962), fit the logarithmic SAR model,
#return a model fit summary and plot the model fit. data(galap) 
fit <- sar_loga(data = galap) 
summary(fit) 
plot(fit)
```

```{r, fig.width=16, fig.height=12}
#Create a fit_collection object containing multiple SAR model fits, and 
#plot all fits. 
fitC <- sar_multi(data = galap, obj = c("power", "loga", "monod"))
plot(fitC) #see Fig.1
```

Model fit validation Model fits can be evaluated through tests of the
normality and homoscedasticity of the residuals. Any of three tests can be
selected to test the normality of the residuals: 1) the Lilliefors extension
of the Kolmogorov normality test (the default; normaTest = “lillie”), 2) the
Shapiro-Wilk test of normality (to be preferred when sample size is small;
“shapiro”), and 3) the Kolmogorov-Smirnov test (“kolmo”). Alternatively, an
option to omit a residuals normality test is provided (“none”). Three options
are provided to check for the homogeneity of the residuals: 1) a correlation
of the residuals with the model fitted values (the default; “cor.fitted”), 2)
a correlation of the residuals with the area values (“cor.area”), or 3) no
homogeneity test (“none”). If a test is selected and is significant at the 5%
level a warning is provided in the model summary; alternatively, the full
results of the three tests can be accessed in the model fit output. A third
model validation check for negative predicted richness values (i.e. when at
least one of the fitted values is negative) is automatically undertaken and a
warning is provided in the model summary if negative values are predicted.

```{r}
#load an example dataset, fit the linear SAR model whilst running residual
#normality and homogeneity tests, and return the results of the residual
#normality test 
data(galap) 
fit <- sar_linear(data = galap, normaTest ="lillie", homoTest = "cor.fitted") 
summary(fit) #a warning is provided  indicating the normality test failed 
fit$normaTest
```

Observed model shape and identifying an asymptote Whilst each of the 20
models has a general shape (@Triantis2012), the actual observed shape of the
model fit can be different, for some models, depending on the parameter
estimates. This is important as the shape of the curve has significant
implications for conservation applications and the testing of macroecological
theory (@Rosenzweig1995). In ‘sars’, the observed shape of a model fit is
determined using the sequential algorithm outlined in @Triantis2012. The
shape is calculated using the model fit within the observed range of area
values. Briefly, the algorithm works by first determining whether the fit is
a straight line. Then, if the fit is classified as not being linear, the
observed shape is classified as either convex or sigmoidal by analysis of the
second derivative (with respect to area) of the model fit (the full algorithm
is detailed in @Triantis2012, p. 220). There has also been considerable
debate in the SAR literature as to whether or not the SAR is asymptotic. In
the ‘sars’ package, to determine whether a fit is asymptotic, for the
relevant models the fitted model parameters are analysed to check whether the
estimated asymptote is within the range of the sample data (@Triantis2012).

Multimodel SAR curve As well as fitting individual models, the package
provides a function (‘sar_average’) to fit up to 20 models, compare the
resultant fits using information criteria, and construct a
multimodel-averaged SAR curve based on information criteria weight (see
@Guilhaumon2010). The multimodel average curve is constructed as a linear
combination of individual model fits by multiplying the predicted richness
values of each of the successfully fitted models by the model’s information
criterion weight, and then summing the resultant values across all models
(@BurnhamAnderson2002). Three information criteria are available in the
package: AIC, AICc and BIC. Confidence intervals around the multimodel
averaged curve can be calculated using a non-parametric bootstrap algorithm
described in @Guilhaumon2010. Briefly, each of the SAR models used in the
‘sar_average’ function is fitted to the data, and the fitted values and
residuals stored. The residuals are then transformed using the approach in
@DavisonHinkley1997 (p.259). For each bootstrap sample, an individual model fit
is selected with the probability of selection being equal to that model’s
information criterion weight. The transformed residuals from this fit are
then sampled with replacement and added to the model’s fitted richness
values. The ‘sar_average’ function is then used to fit all candidate SAR
models to this bootstrapped set of response values, and the multimodel
averaged fitted values stored. Percentile confidence intervals are then
calculated using all bootstrapped fitted values.

The ‘sar_average’ function can be used without specifying any models, in
which case the package attempts to fit each of the 20 models in Table 1;
alternatively, a vector of model names or a ‘fit_collection’ object
(generated using the ‘sar_multi’ function) can be provided using the ‘obj’
argument. The three model validation tests listed above (normality and
homogeneity of residuals, and negative predicted values) can be selected; if
any model fails one or more of the tests during the fitting process it is
removed from the resultant multimodel SAR curve. The output of the
‘sar_average’ function is a list of class ‘multi’ and class ‘sars’, with two
elements. The first element (‘mmi’) contains the multi model inference
(fitted values of the multimodel SAR curve), and the second element
(‘details’) contains a range of information regarding the fitting process,
including the successfully fitted models, the models removed due to failing
any of the validation tests, and the information criterion values, delta
values and weights for the successfully fitted models. The returned object
can easily be plotted using the ‘plot.multi’ generic function, and multiple
plot options are available (using the ‘type’ argument; see Fig. 2). The fits
of all the successfully fitted models and the multimodel SAR curve (with or
without confidence intervals) can be plotted together (‘type’ = “multi”), and
a barplot of the information criterion weights of each model can also be
produced (‘type’ = “bar”).

```{r, fig.width=7, fig.height=19}
#load an example dataset (Niering, 1963), run the ‘sar_average’ function
#using a vector of model names and with no model validation tests, and
#produce the plots in Figure 2 of the paper 
data(niering) 

#run the ‘sar_average’ function using a vector of model names 
fit <- sar_average(data= niering, obj =c("power","loga","koba","mmf","monod",
                                         "negexpo","chapman","weibull3","asymp"),
normaTest = "none", homoTest = "none", neg_check = FALSE, confInt = TRUE, ciN
= 50) #a message is provided indicating that one model (asymp) could not be
#fitted

par(mfrow = c(3,1)) #plot all model fits with the multimodel SAR curve
plot(fit, ModTitle = "a) Multimodel SAR")

#plot the multimodel SAR curve (with confidence intervals; see explanation
#in the main text, above) on its own 
plot(fit, allCurves = FALSE, ModTitle =
      "c) Multimodel SAR with confidence intervals", confInt = TRUE)

#Barplot of the information criterion weights of each model 
plot(fit, type = "bar", ModTitle = "b) Model weights", cex.lab = 1.3)
```

##Additional functions 
In addition to the main functions used to fit and compare the 20 SAR models,
the sars package provides additional functions for specific SAR-based
analyses. First, a function is provided to fit the log–log version of the
power model (a function that is often fitted in SAR studies; Rosenzweig 1995)
and compare parameter values with those generated using the non-linear power
model. The log–log version of the power model is not equivalent to its
non-linear counterpart because of non-equivalence in the study of the
variation in a variable and in its transformation, and bias of
back-transformed results obtained on a logarithmic scale. Second, a function
has been added that enables the fitting of @Coleman1981's random placement
model to a species/sites abundance matrix. According to this model, the
number of species occurring on an island depends on the relative area of the
island and the regional relative species abundances. The fit of the random
placement model can be determined through use of a diagnostic plot (which can
be generated from the function output) of island area (log transformed)
against species richness, alongside the model’s predicted values (see
@Wang2010). Following @Wang2010, the model is rejected if more than a third
of the observed data points fall beyond one standard deviation from the
expected curve. Finally, a function is provided to fit the general dynamic
model of island biogeography (@Whittaker2008) using three different SAR
models (linear, logarithmic and power).

```{r, fig.width=6, fig.height=6}
#load an example dataset, fit the log-log power model, return a model fit
#summary and plot the model fit. When ‘compare’ == TRUE, the non-linear
#power model is also fitted and the resultant parameter values compared. 
#If any islands have zero species, a constant (‘con’) is added to all 
#species richness values. 
data(galap) 
fit <- lin_pow(dat = galap, compare = TRUE, con = 1) 
summary(fit) 
plot(fit)

#load an example dataset, fit the random placement model and plot the 
#model fit and standard deviation. The ‘data’ argument requires a species-
#site abundance matrix: rows are species and columns are sites. The area 
#argument requires a vector of site (island) area values. 
data(cole_sim) 
fit <- coleman(data = cole_sim[[1]], area = cole_sim[[2]]) 
plot(fit, ModTitle = "Hetfield")

#load an example dataset, fit the GDM using the logarithmic SAR model, and
#compare the GDM with three alternative (nested) models: area and time 
#(age of each island), area only, and intercept only. 
data(galap) 
galap$t <- rgamma(16, 5, scale = 2)#add a random time variable 
gdm(data = galap, model = "loga", mod_sel = TRUE)
```

##CONCLUSIONS 
The SAR has been a cornerstone of ecological and biogeographical science for
almost a century and its form and fit are still of great significance in both
theoretical and applied contexts. The development of the ‘sars’ R package
should aid future SAR research by providing a comprehensive set of tools that
enable in-depth exploration of SARs and SAR-related patterns. In addition,
the package has been designed in such a way as to allow other SAR researchers
to add (e.g. via GitHub) new functions and models in the future to ensure the
package is of lasting value. For example, future additions to the package
could include the suite of countryside biogeography SAR models that have
recently been published, standard SAR functions not so far incorporated, or
functions specifically intended for analysis of species accumulation curves
or endemics–area relationships. Finally, whilst the focus of this paper has
been on classic SARs, there is no reason that the functionality in the ‘sars’
package cannot be used to analyse other diversity–area relationships (e.g.
functional or phylogenetic diversity–area relationships). Application of the
full set of 20 models, in addition to the multimodel SAR framework, included
in the ‘sars’ package to a wider range and type of data (e.g. trait and
phylogenetic data) will likely be revealing and will help in improving our
understanding of SARs, and diversity–area relationships more generally.

## References
